---
title: 'Random Walk Metropolis Hastings'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rcpp)
```

We have the state density \(\rho\) for our system of \(N\) particles at temperature \(T\) in a centered cuboid of volume \(\Lambda\). The position of the i-th particle is denoted as \(\mathbf{r}_i \in \mathbb{R}^d \) where \(d\) is the dimension.

\begin{equation}
\pi(\{\mathbf{r}_k\}) \sim \exp\big[ -\frac1T \sum_{i<j} V_{ij} \big]\;, \qquad V_{ij} = \dfrac{q_i q_j}{|\mathbf{r}_i - \mathbf{r}_j|} + \dfrac{1}{|\mathbf{r}_i - \mathbf{r}_j|^8}
\end{equation}

Algorithm: **Random Walk Metropolis-Hastings algorithm**

1: Input: \(X_n\)

2: Generate \(Y_n \sim g(|y - x_n|) \)

3: Set \(X_{n+1}\) according to
\begin{equation}
X_{n+1} =\left\{
                \begin{array}{ll}
                  y_n \;\text{ with probability } \min\left\{\dfrac{\pi(y_n)}{\pi(x_n)},\, 1\right\}\\
                  x_n \;\text{ otherwise}
                \end{array}
              \right.
\end{equation}


We choose our instrumental distribution to be normal gaussian with mean \(\mu = 0\) and width \(\sigma\).
\begin{equation}
g(r) = \dfrac{1}{\sqrt{2\pi\sigma^2}} \exp\left( - \dfrac{r^2}{2\sigma^2} \right)
\end{equation}

The computation of \(\pi(y_n)/\pi(x_n)\) can be simplified with the knowledge that only the vector \(\mathbf{r}_i\) has changed. Only the sum components that depend on the i-th particle have to be considered. Here is \(\mathbf{x}_n = (\mathbf{r}_1, . . ., \mathbf{r}_i, ...,\mathbf{r}_N)\) and \(\mathbf{y}_n = (\mathbf{r}_1, . . ., \mathbf{r}^*_i, ...,\mathbf{r}_N)\) with \(\mathbf{r}^*_i = \mathbf{r}_i + \mathbf{G}\).
\begin{align}
\dfrac{\pi(y_n)}{\pi(x_n)} &= \dfrac{\exp\left[ -\frac1T \sum_{j\neq i} \dfrac{q_i q_j}{|\mathbf{r}^*_i - \mathbf{r}_j|} + \dfrac{1}{|\mathbf{r}^*_i - \mathbf{r}_j|^8} \right]}{\exp\left[ -\frac1T \sum_{j\neq i} \dfrac{q_i q_j}{|\mathbf{r}_i - \mathbf{r}_j|} + \dfrac{1}{|\mathbf{r}_i - \mathbf{r}_j|^8} \right]} \\
&= \exp\left[ -\frac1T \sum_{j\neq i} \dfrac{q_i q_j}{|\mathbf{r}^*_i - \mathbf{r}_j|} - \dfrac{q_i q_j}{|\mathbf{r}_i - \mathbf{r}_j|} + \dfrac{1}{|\mathbf{r}^*_i - \mathbf{r}_j|^8} - \dfrac{1}{|\mathbf{r}_i - \mathbf{r}_j|^8} \right]
\end{align}

```{r visualization, include=FALSE, echo=FALSE}
show_particles <- function(x, q, V){
  l <- V**(1/2)
  # dev.new(width = 550, height = 550, unit = "px")
  par(pty="s")
  plot(NA, xlim=c(-l/2, l/2), ylim=c(-l/2,l/2), xlab='x', ylab='y', asp=1)

  ipos <- which(q==+1)
  posParticles <- x[ipos,]
  # points(x=posParticles[,1],y=posParticles[,2], col='red' )
  arrows(x0=posParticles[,1]-0.5,
         y0=posParticles[,2],
         x1=posParticles[,1]+0.5,
         y1=posParticles[,2],
         length=0.01, angle=90, code=0, col='red')
  arrows(x0=posParticles[,1],
         y0=posParticles[,2]-0.5,
         x1=posParticles[,1],
         y1=posParticles[,2]+0.5,
         length=0.01, angle=90, code=0, col='red')

  ineg <- which(q==-1)
  negParticles <- x[ineg,]
  # points(x=negParticles[,1],y=negParticles[,2], col='blue')
  arrows(x0=negParticles[,1]-0.5,
         y0=negParticles[,2],
         x1=negParticles[,1]+0.5,
         y1=negParticles[,2],
         length=0.01, angle=90, code=0, col='blue')
  arrows(x0=negParticles[,1],
         y0=negParticles[,2]-0.5,
         x1=negParticles[,1],
         y1=negParticles[,2]+0.5,
         length=0.01, angle=90, code=0, col='blue')


  abline(h = -l/2)
  abline(h = +l/2)
  abline(v = -l/2)
  abline(v = +l/2)
}
```

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector timesTwo(NumericVector x) {
  return x * 2;
}
```

```{r}
print(timesTwo(c(1,2)))
```

```{r functions}
# Returns a candidate where the k-th particle was varied.
g <- function(x, k, V, sigma = 1) {
  if (length(x[1,])!=2) print("g is not defined for d!=2")
  
  l <- V**(1/2)
  pi_const <- 3.14159265359
  
  # Generate distance r and direction phi. Get new coord with spherical trafo.
  r <- rnorm(1, mean=0, sd=sigma)
  phi <- runif(1, min=0, max=1)
  x[k,] <- x[k,] + c(cos(phi*pi_const), sin(phi*pi_const)) * r

  # Periodic boundary wrap
  if      (x[k,1] < -l/2) { x[k,1] <- x[k,1] + l }
  else if (x[k,2] < -l/2) { x[k,2] <- x[k,2] + l }
  if      (x[k,1] > +l/2) { x[k,1] <- x[k,1] - l }
  else if (x[k,2] > +l/2) { x[k,2] <- x[k,2] - l }
  
  return(x)
}

# Returns the target distribution probability up to a normalization factor
getE <- function(x, q, t) {
  n <- length(q)     # Number of particles
  d <- length(x[1,]) # Dimensions

  a <- 0 # Coulomb part
  b <- 0 # Hard shell part
  for (i in c(1:(n-1))) {
    for (j in c((i+1):n)) {
      a <- a + q[i] * q[j] / sqrt(sum((x[i,] - x[j,])**2))
      b <- b + 1 / (sqrt(sum((x[i,] - x[j,])**2)))**8
    }
  }
  return((a + b))
}

pi_quot <- function(y, x, q, k, t) {
  n <- length(q)     # Number of particles
  d <- length(x[1,]) # Dimensions

  a <- 0 # Coulomb part
  b <- 0 # Hard shell part
  for (j in c(1:n)[-k]) {
    a <- a + q[k] * q[j] / sqrt(sum((y[k,] - y[j,])**2))
    a <- a - q[k] * q[j] / sqrt(sum((x[k,] - x[j,])**2))
    b <- b + 1 / sum((y[k,] - y[j,])**2)**4
    b <- b - 1 / sum((x[k,] - x[j,])**2)**4
  }
  return((a + b))
}
```

```{r function_random_walk}
random_walk <- function(nIt, nPart, vol, t, X0, q) {
  sigma <- 2   # Stepwidth
  DeltaE <- rep(0,nIt)
  DeltaE[1] <- getE(X0, q, t)

  # Create array for N phasespace points of nPart particles in d dimensions.
  X <- array(data = NA, dim = c(nIt, nPart, d))
  X[1,,] <- X0

  # Execute Markov Chain to make new samples:
  accept_rate <- 0
  for (n in c(2:nIt)) {
    # Generate Y_n
    k <- sample(c(1:nPart), 1)
    y <- g(X[(n-1),,], k, vol, sigma)
    # Set X_n
    DeltaE[n] <- pi_quot(y, X[(n-1),,], q, k, t)
    if (runif(1) < exp(-1/t*DeltaE[n])) {
      X[n,,] <- y
      accept_rate <- accept_rate + 1
    } else {
      DeltaE[n] <- 0
      X[n,,] <- X[(n-1),,]
    }
  }
  # print(accept_rate/nIt*100.)
  return(list(X=X, DeltaE=DeltaE))
}
```

```{r algorithm}
d     <- 2    # Number of dimensions
particleDensity <- 0.1
vol   <- 1e3  # Volume of space
nIt   <- 1e3  # Number of iterations
t     <- 10    # Temperature of heat bath
nPart <- ceiling(vol * particleDensity)  # Number of particles

# Start configuration is chosen uniform randomly in the centered cuboid with volume vol.
# X0 <- array(data = runif(nPart*d, min=-vol**(1/d)/2, max=vol**(1/d)/2), dim = c(nPart, d))
# X0 <- array(data = sample.int(c(1:floor(vol))))
X0 <- array(data = NA, dim = c(nPart, d))
particle_ids <- c(1:nPart)
nRow <- ceiling(sqrt(nPart))
nCol <- ceiling(sqrt(nPart))
for (i in c(1:nPart)) {
  X0[i,1] <- (floor((particle_ids[i]-1) / nCol)) * vol**(1/2) / nRow - vol**(1/2) /2
  X0[i,2] <- ((particle_ids[i]-1) %% nCol      ) * vol**(1/2) / nCol - vol**(1/2) /2
}
# q  <- sample(c(-1,+1), size=nPart, replace=TRUE)
q <- rep(1, nPart)
for (j in c(2:nPart)) {
  q[j] <- q[j-1]*-1
}
# q <- c(rep(+1,ceiling(nPart/2)),rep(-1,floor(nPart/2)))
cat("Number of {positive, negative} particles: {", length(which(q==+1)), ",", length(which(q==-1)),"}\n")

# Show initial distribution
show_particles(X0, q, vol)

Rlist <- random_walk(nIt, nPart, vol, t, X0, q)
DeltaE <- Rlist$DeltaE
X <- Rlist$X

show_particles(X[nIt,,],q, vol)
```
```{r}
E <- cumsum(DeltaE)
plot(c(1:length(DeltaE)), cumsum(DeltaE))
acf(E[c((nIt-1000):nIt)])

hist(E[-c(1:5000)])
```
```{r}
library(LaplacesDemon)
Temp <- rep(0, 20)
for (k in c(1:length(Temp))) {
  Temp[k] <- 2/k
}
iats <- rep(0, length(Temp))
for (i in c(1:length(Temp))) {
  cat("Temp = ", Temp[i], " \n")
  Reslist <- random_walk(1e4, nPart, 1e3, Temp[i], X0, q)
  iats[i] <- IAT(Reslist$DeltaE[-c(1:5000)])
}
```
```{r}
plot(1/Temp, iats, ylab = "IAT")
```

```{r analysis_minimalDistance}
hist_minDists <- rep(Inf, nIt)
for (n in seq(from=1, to=nIt, by=ceiling(nIt/60))) {
  for (i in c(1:(nPart-1))) {
    for (j in c((i+1):nPart)) {
      d <- sqrt(sum((X[n,i,] - X[n,j,])**2))
      if (d < hist_minDists[n]) hist_minDists[n] <- d
    }
  }
}
hist_minDists[which(hist_minDists==Inf)] <- NA
```
```{r plot_minimalDistance}
idx <- !is.na(hist_minDists)
plot(x=c(1:nIt)[idx] ,y=hist_minDists[idx], type='b', log='y', ylim=c(0.01,2))
```

