---
title: 'Steering File for the MCMC for Particles in a Potential Analysis'
output: html_document
---

This steering script is used to analyse the particles in a potential system. It is intended to be used in a notebook style. Knitting the .html does not work currently for us, because it does not compile Rcpp code correctly.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Rcpp)

# Source helper functions
source("Helper.R")
# Source R functions related to the Markov chain
source("MCMC_functions.R")
# Source Rcpp functions related to the Markov chain. Mind the correct path!
sourceCpp("C:/Ubuntu Subsystem Folder/particles-in-a-potential/MCMC_functions.cpp")
```

Simulate a single run of the Markov chain and evaluate it
```{r MCMC_Simulation_Single}
# --------------------------------------------------------------------------------
# Only change these parameters:
# EXTERIOR PARAMETERS:
dim             <- 2     # Number of dimensions
nPart           <- 49    # Number of simulated particles
rho             <- 0.05  # Particle Density
t               <- 3e-2  # Temperature of heat bath
# SIMULATION PARAMETERS:
nIt             <- 2e6   # Number of Markov chain iterations
start_type      <- "ordered" # Start configuration type (ordered / random)
# Other parameters:
showConfigBefore<- TRUE
showConfigAfter <- TRUE
saveConfigAfter <- FALSE
saveConfigAfterAs <- "configuration.pdf"
# --------------------------------------------------------------------------------



# Do not touch below here.
# --------------------------------------------------------------------------------
# Generate start configuration: (type="ordered" or type="random")
init_config <- gen_initConfig(type=start_type, dim=dim, vol=nPart/rho, nPart=nPart)
# --------------------------------------------------------------------------------
if (showConfigBefore) show_particles(init_config$X0, init_config$q, nPart/rho)
timer.start <- Sys.time()
# Start the MCMC simulation
MCMC_Single_Run <- MCMC_CPP(nIt, nPart, nPart/rho, t, sigma=1, N_burnIn=0, init_config$X0, init_config$q)
timer.end   <- Sys.time()
if (showConfigAfter) {
  if (saveConfigAfter) pdf("config_rho010_t0001.pdf", width=10*0.55, height=10*0.55)
  show_particles(MCMC_Single_Run$X, init_config$q, nPart/rho)
  if (saveConfigAfter) dev.off()
}
# --------------------------------------------------------------------------------
# Output of some observables:
cat("Number of {positive, negative} particles: {", length(which(init_config$q==+1)), ",", length(which(init_config$q==-1)),"}\n")
cat("Runtime: ",as.double(timer.end - timer.start, units='secs'),"\n")
cat("Acceptrate: ", MCMC_Single_Run$Accpetance*100,"%\n")
cat("Average cluster size: q= ", mean(MCMC_Single_Run$q), "\n")
cat("Average minimal distance to nearest neighbour: Omega= ", mean(MCMC_Single_Run$omega), "\n")
cat("Average distance between particles: Lambda= ", mean(MCMC_Single_Run$lambda), "\n")
cat("Average number of direct neighbours: Kappa= ", mean(MCMC_Single_Run$kappa), "\n")
# --------------------------------------------------------------------------------
rm(init_config, dim, nPart, rho, saveConfigAfter, saveConfigAfterAs, showConfigAfter, showConfigBefore, start_type, t, timer.start, timer.end)
# --------------------------------------------------------------------------------
```

```{r MCMC_Simulation_Single_Evaluation}
E <- MCMC_Single_Run$E
x.ticks <- c(1e1:nIt)
plot(x=x.ticks,y=E[x.ticks],type='l',log='')
rm(x.ticks)

acf_range <- c((nIt-1e5):nIt)
acf(E[acf_range], lag.max = 50000)

cat("IAT = ",IAT_CPP(E[acf_range]),"\n")
rm(acf_range)
```



Create a new dataframe which will store the results of multiple runs and which will be written to the hard drive.
```{r}
file_name <- "MCMC_Results.RData"

# frameHelper.createNewFrame(file_name) # BE CAREFULL NOT TO ACCIDENTALLY OVERWRITE EXISTING FILE

frameHelper.showContentOverview(file_name)
```

This code chunk simulates nResamples many Markov Chain Monte Carlo runs for length(x.T) many temperaturs and saves it in the dataframe to disk.
```{r MCMC_Swipe}
# --------------------------------------------------------------------------------
# LOAD RESULTS DATA FRAME FROM DISK
load(file_name)
run_no <- max(resFrame$RUN_ID, na.rm=TRUE)+1
# --------------------------------------------------------------------------------
# Standard Sets:
# 1D: nPart <- 50, rho <- 0.10
# 2D: nPart <- 50, rho <- 0.05
# 3D: nPart <- 64, rho <- 0.01
# EXTERIOR PARAMETERS:
dim             <- 2
nPart           <- 50
rho             <- 0.05
x.T <- exp(log(10)*seq(from=-3.2, to=1.2, by=0.2))
x.T <- c(x.T, exp(log(10)*seq(from=-2.1, to=-0.3, by=0.05))) # More detail in transition region
# SIMULATION PARAMETERS
nIt             <- 0.5e6
nResamples      <- 5
sct             <- "ordered"
# --------------------------------------------------------------------------------





# --------------------------------------------------------------------------------
lowerBound <- 1e-2
upperBound <- 1e-1
# x.T <- x.T[which(x.T>=lowerBound)]
# x.T <- x.T[which(x.T< upperBound)]
x.T <- sort(x.T)
x.T <- x.T[!duplicated(round(x.T,8))]

x.T <- x.T[which(c(1:length(x.T))%%2==0)] # Each of those reduce the amount
x.T <- x.T[which(c(1:length(x.T))%%2==1)] # of simulated temperatures in
# x.T <- x.T[which(c(1:length(x.T))%%2==0)] # half for more rough swipes

# --------------------------------------------------------------------------------

resFrame <- myMCMCSwipe(dim=dim, nPart=nPart, rho=rho, sigma=1, sct=sct, nIt=nIt, nResamples=nResamples, x.beta=1/x.T, df=resFrame)

save(resFrame, file=file_name)
rm(resFrame)

rm(run_no, dim, nPart, rho, nIt, nResamples, x.T, lowerBound, upperBound, sct)
```

This code chunk loads the data for the requested exterior parameters from the given dataframe and plots the data for the q, kappa parameters (as well as the omega and lambda parameters).
```{r plot_orderParams}
load(file_name)
# --------------------------------------------------------------------------------
# EXTERIOR PARAMETERS:
dim             <- 2
nPart           <- 50
rho             <- 0.05
# --------------------------------------------------------------------------------
x.beta <- frameHelper.getBetaValues(dim=dim, vol=nPart/rho, rho=rho)
# x.beta <- x.beta[which(x.beta>3e0)]

plt.data <- frameHelper.getObservables(dim=dim, vol=nPart/rho, rho=rho, beta=x.beta)
rm(x.beta)

myPlotData(1/plt.data$x, plt.data$y.q, plt.data$y.q_Err)
abline(h=1)
myPlotData(1/plt.data$x, plt.data$y.lambda, plt.data$y.lambda_Err)
myPlotData(1/plt.data$x, plt.data$y.omega , plt.data$y.omega_Err )
myPlotData(1/plt.data$x, plt.data$y.kappa , plt.data$y.kappa_Err )

rm(resFrame)
rm(plt.data, dim, rho)
```

This chunk generates the q+kappa plots and fits a function to the datapoints.
```{r q_kappa_plots_with_fit}
load(file_name)
# --------------------------------------------------------------------------------
dim             <- 2
nPart           <- 50
rho             <- 0.05
# --------------------------------------------------------------------------------
x.beta <- frameHelper.getBetaValues(dim=dim, vol=nPart/rho, rho=rho)

plt.data <- frameHelper.getObservables(dim=dim, vol=nPart/rho, rho=rho, beta=x.beta)
rm(x.beta)

#plot custom
x <- 1/plt.data$x     
yq    <- plt.data$y.q         * 0.1
yqerr <- plt.data$y.q_Err     * 0.1
yk    <- plt.data$y.kappa     * 1.0
ykerr <- plt.data$y.kappa_Err * 1.0


xlim <- c(1e-5, 1e5)
xlim <- c(1e-3, 1e0)
ylim <- c(0, max(yq+yqerr, yk+ykerr, na.rm=TRUE))

plot(NA, NA, xlim=xlim, ylim=ylim, type = "l", log="x", main = "", xlab = "T", ylab = "Value")

draw.DataWithError(x=x, y=yq, dy=yqerr, col="blue", bars=TRUE, points=TRUE, lines=FALSE) #q
draw.DataWithError(x=x, y=yk, dy=ykerr, col="red"  , bars=TRUE, points=TRUE, lines=FALSE) #kappa

legend("topright", legend = c("q", expression(kappa)), lty = c(1,1), col = c("blue", "red"))

# Fit of function to the data points:
# Fit to q: (initial parameters may have to be adjusted to guarantee convergence)
parsq <- fit.automaticRoutine(startparam=c(6.4, 0.013, 0.1, 0.035), x=x, y=yq, dy=yqerr, xrange=c(1e-3,1e2), fn=chisqr_PiecewiseLinAndExpFunction)
fit.draw(parsq$par, x=x, col="black", lines=TRUE, vlines=FALSE)

# Fit to kappa:
parsk <- fit.automaticRoutine(startparam=c(3, 0.005, 0.1, 0.5), x=x, y=yk, dy=ykerr, xrange=c(1e-3,1e2), fn=chisqr_PiecewiseLinAndExpFunction)
fit.draw(parsk$par, x=x, col="black", lines=TRUE, vlines=FALSE)

print_fitRes <- function(varname, par, parErr) {
  cat("\nFit to ",varname,": \n")
  cat("A1:        ", par$par[1], "\t +- " , parErr$A1err, "\n")
  cat("x1:        ", par$par[2], "\t +- " , parErr$X1err, "\n")
  cat("A2:        ", par$par[3], "\t +- " , parErr$A2err, "\n")
  cat("x2:        ", par$par[4], "\t +- " , parErr$X2err, "\n")
  cat("Chi^2/dof: ", par$chi2/par$dof  , "\t +- " , parErr$chi2err,  "\t , p-value: ", par$pval, "\n")
}
print_fitRes("q",parsq$par, parsq$parerr)
print_fitRes("kappa",parsk$par, parsk$parerr)
rm(print_fitRes)

rm(resFrame)
rm(x, xlim, ylim, yq, yqerr, yk, ykerr)
rm(plt.data, dim, rho, nPart)
rm(parsq, parsk)
```


The burn-in is analyzed in the following three chunks.
```{r burnin_energy_setup}
# --------------------------------------------------------------------------------
# EXTERIOR PARAMETERS:
dim   <- 2    # Number of dimensions
nPart <- 50
rho   <- 0.05
# SIMULATION PARAMETERS:
nIt   <- 1e6  # Number of iterations
type <- "random"


t     <- 5e-2 
init_config <- gen_initConfig(type=type, dim=dim, vol=nPart/rho, nPart=nPart)
MCMC_BurnIn_Run1 <- MCMC_CPP(nIt, nPart, nPart/rho, t, sigma=1, N_burnIn=0, init_config$X0, init_config$q)

t     <- 1e-3 
init_config <- gen_initConfig(type=type, dim=dim, vol=nPart/rho, nPart=nPart)
MCMC_BurnIn_Run2 <- MCMC_CPP(nIt, nPart, nPart/rho, t, sigma=1, N_burnIn=0, init_config$X0, init_config$q)

# show_particles(MCMC_BurnIn_Run$X, init_config$q, vol)
rm(dim, rho, t, nPart, init_config, type)
# --------------------------------------------------------------------------------
```

```{r burnin_energy_plot}
nIt <- 1e5
x.ticks <- c(1:nIt)

E1 <- MCMC_BurnIn_Run1$E
E2 <- MCMC_BurnIn_Run2$E

# pdf("burnin_energy.pdf", width=10*0.7, height=6*0.7)
plot(NA, NA, xlim=c(-5e2,nIt), ylim=c(-21,-15), xlab="Markov Chain Step", ylab="Energy", log="", xaxs="i", yaxs="r")
lines(x.ticks, E1[x.ticks], col="red")
lines(x.ticks, E2[x.ticks], col="blue")

legend("topright", legend = c("T = 0.050", "T = 0.001"), lty = c(1,1), col = c("red", "blue"))
options(scipen=10)

# dev.off()
rm(x.ticks)
```

```{r burnin_energy_acf}
nInt             <- 1e5
lag.max          <- 10000
acf_range <- c((nIt-nInt):nIt)

E1 <- MCMC_BurnIn_Run1$E
acf1 <- acf(E1[acf_range], lag.max = lag.max, plot=TRUE)

E2 <- MCMC_BurnIn_Run2$E
acf2 <- acf(E2[acf_range], lag.max = lag.max, plot=TRUE)

rm(nInt, acf_range)
```